---
title: "PCAEuropeanCountries"
author: "Duc-Anh Nguyen"
date: "2025-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## EDA
```{r}
euro<-read.table("europa.txt",header=T)
dim(euro)
head(euro)
skimr::skim(euro)


#asnumericmatrix
euro_matrix <- euro[, -1] |>
  as.matrix()
rownames(euro_matrix) <- euro[, 1]
head(euro_matrix)

(cov_euro <- cov(euro_matrix))
(cor_euro <- cor(euro_matrix))
```

Let's do PCA with SVD   

##PCA
```{r}
pca_corr <- prcomp(euro_matrix, scale = TRUE)
pca_corr
summary(pca_corr)
```

let us visualise that   
```{r}
#install.packages("cowplot")
library(cowplot)
library(factoextra)

screeplot(pca_corr)

# plot the scree plot
plot_grid(
  fviz_eig(pca_corr, addlabels = TRUE) + ggtitle(""),
  fviz_eig(pca_corr, choice = "eigenvalue", addlabels = TRUE) + ggtitle("")
)
```

Consider the Kaiser criteria (we keep all the PCs that has eigenvalue larger than 1), the variance criteria(all the PCs that explain 80% the variance) and the practicality, we will stick with 2 PCs


Now for biplot 
```{r}
fviz_pca_var(pca_corr, col.var = "black")

fviz_pca_ind(pca_corr)
```
arbl(Unemployment) and brut(Gross national product) have negative correlation, while ober(country area) and einw(population) are strong positive correlated. Between arbl and ober/einw there is pretty much no correlation, same for brut and ober/einw.

From the biplot we can see for example that Germany is a country with large living area and population. Norway is a country having high gross national product and low rate of unemployment, while Poland is the opposite.

Note that we could get the coordinates of each country with
```{r}
pca_corr$x[, 1:2]
```
